Miniproject - Analytic Number Theory

### 2021-12-16 Analytic number theory/submission.tex
Improved display of mod in ANT

For convenience, we include the definition of a Dirichlet character from the task description first.
\begin{definition}
    Let $q \geq 2$, then a \emph{Dirichlet character (mod $q$)} is a function $\chi: \N \to \C$ such that
    \begin{itemize}
        \item $\chi$ is completely multiplicative, so $\chi(a)\chi(b) = \chi(ab)$
        \item $\chi$ is periodic modulo $q$, so $\chi(n + q) = \chi(n)$
        \item $\chi(n) \neq 0$ if and only if $n \perp q$
    \end{itemize}
\end{definition}
First, we will give another characterization of Dirichlet characters.
\begin{lemma}[Characterization of Dirichlet characters]
    \label{prop:characterization_dirichlet_character}
    We have a one-to-one correspondence between Dirichlet characters mod $q$ and group homomorphisms $(\Z/q\Z)^\units \to \C^\units$ via
    \begin{align*}
        \{ f: (\Z/q\Z)^\units \to \C^\units \ | \ f \ \text{group hom} \} &\to \{ \chi: \N \to \C \ | \ \chi \ \text{Dirichlet character mod $q$} \} \\
        f &\mapsto \chi_f := \left( \N \to \C, \ n \mapsto \begin{cases}
            f([n]) & \text{if $n \perp q$} \\
            0 & \text{otherwise}
        \end{cases} \right)
    \end{align*}
\end{lemma}
\begin{proof}
    First of all, we show that the map is well-defined. Let $f: (\Z/q\Z)^\units \to \C^\units$ be a (multiplicative) group homomorphism, and we show that $\chi_f$ is a Dirichlet character.

    Note that property (ii) and (iii) directly follow from the definition, as $\chi_f(n)$ only depends on the value of $n \mod q$.
    So consider some $a, b \in \N$.
    If both $a \perp q$ and $b \perp q$ then
    \begin{equation*}
        \chi_f(a)\chi_f(b) = \chi([a])\chi([b]) = \chi([ab]) = \chi_f(ab)
    \end{equation*}
    as also $ab \perp q$.

@@

            0 & \text{if $n \equiv 0, 2 \mod 4$} \\
            1 & \text{if $n \equiv 1 \mod 4$} \\
            -1 & \text{if $n \equiv 3 \mod 4$}

@@

    (this is a group homomorphism, as $3^2 = 9 \equiv 1 \mod 4$)

@@

    Let $n \equiv 3 \mod 4$. Then $n$ has a prime factor $p \equiv 3 \mod 4$.

@@

    However, $3 \equiv n$ is not a square modulo $4$, so find that $a \not\equiv b \mod 4$.
    As both $a$ and $b$ must be odd, we see that either $a \equiv 3 \mod 4$ or $b \equiv 3 \mod 4$ and the claim follows by the induction hypothesis.

@@

    There are infinitely many primes $p$ with $p \equiv 3 \mod 4$.

@@

        P \equiv 3^{2 \lceil \frac N 2 \rceil} \equiv 1^{\lceil \frac N 2 \rceil} = 1 \mod 4

@@

    Thus, by Lemma~\ref{prop:prime_factor_3_mod_4}, $P + 2$ has a prime factor $q \equiv 3 \mod 4$.

@@

For the case of primes $\equiv 1 \mod 4$, I have remembered Fermat's theorem on the sum of two squares and its connection to primes in the ring $\Z[i]$ of Gaussian integers, and somehow my train of thoughts went into Algebraic Number Theory.

@@

    Then there are infinitely many primes $p$ with $p \equiv 1 \mod q$.

@@

    Let now $K = \Q(\omega_q)$ be the $q$-th cyclotomic number field with a primitive $q$-th root of unity $\omega_q$ (i.e. $\Phi_q(\omega_q) = 0$).
    Let further $\mathcal{O} \subseteq K$ be the ring of integral elements over $\Z$ in $K$.
    The prime decomposition law for Dedekind ring extension \cite[Chapter I, Prop 8.3]{neukirch} tells us that for a prime $p$, the ideal $(p)$ is reducible in $\mathcal{O}$ if and only if $\Phi_q \mod p$ is reducible.
    As $(\Z/p\Z)^\units$ is cyclic of order $p - 1$, this is the case if and only if $q \divides p - 1$, i.e. $p \equiv 1 \mod q$.

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
In a later chapter, the book also proves the general result \cite[Chapter VII, Theorem 5.14]{neukirch}.
I am pretty sure that those are exactly the functions $L(\chi, s)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \mod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \mod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \mod 4$ and 332398 primes $\equiv 3 \mod 4$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

    It is a fact from Algebra that $S_q \cong (\Z/q\Z)^\units$, hence Dirichlet characters modulo a prime $q$ are in 1-to-1 correspondence with the endomorphisms $\mathrm{End}((\Z/q\Z)^\units)$ of $(\Z/q\Z)^\units$.
\end{proposition}
\begin{proof}
    We have that $S_q = \{ x \in \C^\units \ | \ x^{q - 1} = 1 \}$ and the claim directly follows from Corollary~\ref{prop:dirichlet_character_image_group}.
\end{proof}
Note that the endomorphisms $\mathrm{End}((\Z/q\Z)^\units)$ form a ring w.r.t. pointwise multiplication and composition
(this is the case for the endomorphisms of any abelian group).
In particular, the endomorphisms and so the Dirichlet characters form a group w.r.t. pointwise multiplication.
\begin{definition}
    Denote by $\mathrm{Dir}(q)$ the set of Dirichlet characters modulo $q$.
\end{definition}
By Corollary~\ref{prop:dirichlet_character_image_group} each Dirichlet character resp. group endomorphism of $(\Z/q\Z)^\units$ is determined by its value at a primitive root of unity $r \in (\Z/q\Z)^\units$.
Further, by Corollary~\ref{prop:dirichlet_character_primitive_root}, we also have a Dirichlet character resp. group endomorphism for each possible value at a primitive root of unity.
Hence
\begin{equation*}
    |\mathrm{Dir}(q)| = |\mathrm{End}((\Z/q\Z)^\units)| = |(\Z/q\Z)^\units| = q - 1
\end{equation*}
It follows that there are exactly $q - 1$ distinct Dirichlet characters modulo a prime $q$.
\begin{remark}
    It is again a fact that $(\Z/p^k\Z)^\units$ is cyclic for an odd prime $p$ and $k \geq 1$.
    Hence, everything up to now can also be done for odd prime powers, if we replace $q - 1$ by $\phi(q)$.
\end{remark}
Because of Lemma~\ref{prop:dirichlet_series_at_one} it might seem like a good idea to study in which cases the value $\sum_{x \in (\Z/q\Z)^\units} \chi(x)$ is zero.
\begin{proposition}[Question (iii)]
    \label{prop:finite_sums_characters}
    Let $\chi_0$ be the trivial Dirichlet character given by $r \mapsto 1$.
    Then
    \begin{align*}
        \sum_{a \in (\Z/q\Z)^\units} \chi(a) &= \begin{cases}
            q - 1 & \text{if $\chi = \chi_0$} \\
            0 & \text{otherwise}
        \end{cases}, \\
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a) &= \begin{cases}
            q - 1 & \text{if $a \equiv 1 \mod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{align*}
    Furthermore, for $b \perp q$ have
    \begin{equation*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a)\overline{\chi(b)} = \begin{cases}
            q - 1 & \text{if $a \equiv b \mod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
\end{proposition}
\begin{proof}
    Clearly
    \begin{equation*}
        \sum_{a \in (\Z/q\Z)^\units} \chi_0(a) = q - 1 \quad \text{and} \quad \sum_{\chi \in \mathrm{Dir}(q)} \chi(1) = \sum_{\chi \in \mathrm{Dir}(q)} 1 = q - 1
    \end{equation*}
    So it is left to show that we get zero in the other cases.

    Consider a Dirichlet character $\chi \neq \chi_0$ given by $r \mapsto \xi$ for a primitive $q$-th root of unity $\xi \neq 1$.
    Then
    \begin{equation*}
        \sum_{a \in (\Z/q\Z)^\units} \chi(a) = \sum_{k = 0}^{q - 2} \chi(r^k) = \sum_{k = 0}^{q - 2} \xi^k = \frac {1 - \xi^{q - 1}} {q - \xi} = 0 
    \end{equation*}
    By using the earlier results on the structure of $\mathrm{Dir}(q)$ we see that for $a \equiv r^k \not\equiv 1 \mod q$, have
    \begin{align*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a) &= \sum_{\chi \in \mathrm{Dir}(q)} \chi(r)^k \\
        &= \sum_{\xi \ \text{$q$-th root of unity}} \xi^k = \sum_{l = 0}^{q - 2} \omega^{kl} = \frac {1 - (\omega^{q - 1})^k} {1 - \omega^k} = 0 
    \end{align*}
    where $\omega \in \C$ is any primitive $q$-th root of unity and $r \in \Z$ is a primitive root modulo $q$.

    For the last part, note that for any $q$-th root of unity $\xi$, we have $\xi\overline{\xi} \in \R$ with $\xi\overline{\xi} = |\xi|^2 > 0$.
    Furthermore, $\overline{\xi}$ is also a $q$-th root of unity, and so we see that $\xi\overline{\xi} = 1$ (the only real, positive root of unity is 1).
    It follows that for any Dirichlet character $\chi$ have $\overline{\chi([a])} = \chi([a]^{-1})$.
    Thus
    \begin{equation*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a)\overline{\chi(b)} = \sum_{\chi \in \mathrm{Dir}(q)} \chi([a][b]^{-1}) = \begin{cases}
            q - 1 & \text{if $[a][b]^{-1} = 1 \in (\Z/q\Z)^\units$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    The condition $ab^{-1} = 1$ is equivalent to $a \equiv b \mod q$, so the claim follows.
\end{proof}
Using these basic results, we can now prove facts on the Dirichlet series of characters.
\begin{proposition}[Question (iv)]
    \label{prop:representation_mangoldt_series_congruent_a}
    Let $a \perp q$. Then for $\Re(s) > 1$ have
    \begin{equation*}
        \sum_{n \equiv a \mod q} \frac {\Lambda(n)} {n^s} = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
    (All those series obviously converge absolutely since $\Re(s) > 1$)
\end{proposition}
\begin{proof}
    By Prop.~\ref{prop:finite_sums_characters} we have for all $n \in \N$ that
    \begin{equation*}
        \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \chi(n) = \begin{cases}
            1 & \text{if $a \equiv n \mod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    It follows that
    \begin{align*}
        \sum_{n \equiv a \mod q} \Lambda(n) n^{-s} =& \sum_{n \geq 1} \Lambda(n) n^{-s} \begin{cases}
            1 & \text{if $a \equiv n \mod q$} \\
            0 & \text{otherwise}
        \end{cases} \\
        =& \sum_{n \geq 1} \Lambda(n) n^{-s} \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \chi(n) \\
        =& \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \Lambda(n) \chi(n) n^{-s} 
    \end{align*}
    as infinite summation clearly commutes with finite sums.
\end{proof}
\begin{example}[Question (v)]
    \label{ex:nontrivial_dirichlet_characters_mod_5}
    We consider the Dirichlet characters mod 5.
    A (primitive) 5-th root of unity $\omega_5 \in \C$ is given by $\omega_5 = \exp(2\pi i/5)$.
    On the other hand, a primitive root modulo 5 is e.g. given by $r = 2$ since $2^2 \equiv -1 \mod 5$.
    Thus we have the trivial Dirichlet character $\chi_0$ and $5 - 1 = 4$ nontrivial Dirichlet characters mod 5, namely those given by
    \begin{align*}
        \chi_1: 1 &\mapsto 1,\ 2 \mapsto \omega_5 = \exp(2\pi i/5),\ 3 \mapsto \omega_5^3 = \exp(6\pi i/5),\ 4 \mapsto \omega_5^2 = \exp(4\pi i/5),\ \\
        \chi_2: 1 &\mapsto 1,\ 2 \mapsto \omega_5^2 = \exp(4\pi i/5),\ 3 \mapsto \omega_5 = \exp(2\pi i/5),\ 4 \mapsto \omega_5^4 = \exp(8\pi i/5),\ \\
        \chi_3: 1 &\mapsto 1,\ 2 \mapsto \omega_5^3 = \exp(6\pi i/5),\ 3 \mapsto \omega_5^4 = \exp(8\pi i/5),\ 4 \mapsto \omega_5^1 = \exp(2\pi i/5),\ \\
        \chi_4: 1 &\mapsto 1,\ 2 \mapsto \omega_5^4 = \exp(8\pi i/5),\ 3 \mapsto \omega_5^2 = \exp(4\pi i/5),\ 4 \mapsto \omega_5^3 = \exp(6\pi i/5),\ \\
    \end{align*}
\end{example}

@@

    More concretely, consider
    \begin{equation*}
        f: \N \to \C, \quad n \mapsto \begin{cases}
            1 & \text{if $n \equiv a \mod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    Note that this function is not multiplicative.
    After applying the first step, we are interested in the poles of the meromorphic function
    \begin{equation*}
        \sum_{n \geq 1} f(n) \Lambda(n) n^{-s}
    \end{equation*}
    I will not prove this here, but via partial summation, we can find a meromorphic continuation to $\Re(s) > 0$.

>>>>

For convenience, we include the definition of a Dirichlet character from the task description first.
\begin{definition}
    Let $q \geq 2$, then a \emph{Dirichlet character (modulo $q$)} is a function $\chi: \N \to \C$ such that
    \begin{itemize}
        \item $\chi$ is completely multiplicative, so $\chi(a)\chi(b) = \chi(ab)$
        \item $\chi$ is periodic modulo $q$, so $\chi(n + q) = \chi(n)$
        \item $\chi(n) \neq 0$ if and only if $n \perp q$
    \end{itemize}
\end{definition}
First, we will give another characterization of Dirichlet characters.
\begin{lemma}[Characterization of Dirichlet characters]
    \label{prop:characterization_dirichlet_character}
    We have a one-to-one correspondence between Dirichlet characters modulo $q$ and group homomorphisms $(\Z/q\Z)^\units \to \C^\units$ via
    \begin{align*}
        \{ f: (\Z/q\Z)^\units \to \C^\units \ | \ f \ \text{group hom} \} &\to \{ \chi: \N \to \C \ | \ \chi \ \text{Dirichlet character mod $q$} \} \\
        f &\mapsto \chi_f := \left( \N \to \C, \ n \mapsto \begin{cases}
            f([n]) & \text{if $n \perp q$} \\
            0 & \text{otherwise}
        \end{cases} \right)
    \end{align*}
\end{lemma}
\begin{proof}
    First of all, we show that the map is well-defined. Let $f: (\Z/q\Z)^\units \to \C^\units$ be a (multiplicative) group homomorphism, and we show that $\chi_f$ is a Dirichlet character.

    Note that property (ii) and (iii) directly follow from the definition, as $\chi_f(n)$ only depends on the value of $n$ modulo $q$.
    So consider some $a, b \in \N$.
    If both $a \perp q$ and $b \perp q$ then
    \begin{equation*}
        \chi_f(a)\chi_f(b) = \chi([a])\chi([b]) = \chi([ab]) = \chi_f(ab)
    \end{equation*}
    as also $ab \perp q$.

@@

            0 & \text{if $n \equiv 0, 2 \pmod 4$} \\
            1 & \text{if $n \equiv 1 \pmod 4$} \\
            -1 & \text{if $n \equiv 3 \pmod 4$}

@@

    (this is a group homomorphism, as $3^2 = 9 \equiv 1 \pmod 4$)

@@

    Let $n \equiv 3 \pmod 4$. Then $n$ has a prime factor $p \equiv 3 \pmod 4$.

@@

    However, $3 \equiv n$ is not a square modulo $4$, so find that $a \not\equiv b \pmod 4$.
    As both $a$ and $b$ must be odd, we see that either $a \equiv 3$ or $b \equiv 3 \pmod 4$ and the claim follows by the induction hypothesis.

@@

    There are infinitely many primes $p$ with $p \equiv 3 \pmod 4$.

@@

        P \equiv 3^{2 \lceil \frac N 2 \rceil} \equiv 1^{\lceil \frac N 2 \rceil} = 1 \pmod 4

@@

    Thus, by Lemma~\ref{prop:prime_factor_3_mod_4}, $P + 2$ has a prime factor $q \equiv 3 \pmod 4$.

@@

For the case of primes $\equiv 1 \pmod 4$, I have remembered Fermat's theorem on the sum of two squares and its connection to primes in the ring $\Z[i]$ of Gaussian integers, and somehow my train of thoughts went into Algebraic Number Theory.

@@

    Then there are infinitely many primes $p$ with $p \equiv 1 \pmod q$.

@@

    Let now $K = \Q(\omega_q)$ be the $q$-th cyclotomic number field with a primitive $q$-th root of unity $\omega_q$ (i.e. $\Phi_q(\omega_q) = 0$).
    Let further $\mathcal{O} \subseteq K$ be the ring of integral elements over $\Z$ in $K$.
    The prime decomposition law for Dedekind ring extension \cite[Chapter I, Prop 8.3]{neukirch} tells us that for a prime $p$, the ideal $(p)$ is reducible in $\mathcal{O}$ if and only if $\Phi_q$ is reducible modulo $p$.
    As $(\Z/p\Z)^\units$ is cyclic of order $p - 1$, this is the case if and only if $q \divides p - 1$, i.e. $p \equiv 1 \pmod q$.

    Now consider the element $\alpha = \omega_q - xqP \in \mathcal{O}$. Then
    \begin{align*}
        N_{K/\Q}(\alpha) &= \prod_{\sigma: K \to \C \ \text{$\Q$-field homomorphism}} \sigma(\omega_q - xqP) \\
        &= \prod_{\sigma} (\sigma(\omega_q) - xqP) = \mathrm{MiPo}_\Q(\omega_q)(xqP) = \Phi_q(xqP) \neq 1
    \end{align*}
    as $\mathrm{MiPo}_\Q(\omega_q) = \prod_\sigma (\sigma(\omega_q) - X)$.
    Hence, $\alpha$ is not a unit in $\mathcal{O}$.
    On the other hand, $(\alpha)$ is coprime to $(p_i)$ for each $p_i$, as 
    \begin{equation*}
        \omega_q = \alpha - xqP \in (\alpha) + (p_i) \quad \text{and} \quad \omega_q \in \mathcal{O}^\units
    \end{equation*}
    By our assumption, the only prime ideals in $\mathcal{O}$ are the prime ideal factors of $(p_i)$ and $(p)$ for $p \neq p_i$.
    Thus, the prime ideal factorization of $(\alpha)$ consists only of prime ideals $(p), p \neq p_i$ and it follows that $(\alpha) = (n)$ for some integer $n \geq 2$.
    As $\omega_q$ and $xqP \in \Z$ are $\Q$-linearly independent, we see that $n \divides \omega_q$ and $n \divides xqP$.
    However, the former is a contradiction, as $\omega_q \in \mathcal{O}^\units$ is a unit and no $n \geq 2$ is a unit.
\end{proof}
In a later chapter, the book also proves the general result \cite[Chapter VII, Theorem 5.14]{neukirch}.
I am pretty sure that those are exactly the functions $L(\chi, s)$ that we defined above.
\begin{corollary}[Question (iii)]
    There are infinitely many primes $p$ with $p \equiv 1 \pmod 4$.
\end{corollary}
\begin{proof}
    This is just a special case of Prop.~\ref{prop:primes_1_mod_q}.
\end{proof}
\begin{example}[Question (iii)]
    Using a computer, we can also study the actual frequency of prime numbers $\equiv 1, 3 \pmod 4$ among e.g. the first $10^8$ integers.
    This seems to indicate that both numbers are asymptotically equal, which seems natural, given the result of Example~\ref{ex:dirichlet_sum_over_primes}.
    For example, there are 332180 primes $\equiv 1 \pmod 4$ and 332398 primes $\equiv 3$ smaller than $10^8$.
    To find these numbers, the following python code was used.
    \lstinputlisting[language = python]{./primes.py}
\end{example}

@@

    It is a fact from Algebra that $S_q \cong (\Z/q\Z)^\units$, hence Dirichlet characters modulo a prime $q$ are in 1-to-1 correspondence with the endomorphisms $\mathrm{End}((\Z/q\Z)^\units)$ of $(\Z/q\Z)^\units$.
\end{proposition}
\begin{proof}
    We have that $S_q = \{ x \in \C^\units \ | \ x^{q - 1} = 1 \}$ and the claim directly follows from Corollary~\ref{prop:dirichlet_character_image_group}.
\end{proof}
Note that the endomorphisms $\mathrm{End}((\Z/q\Z)^\units)$ form a ring w.r.t. pointwise multiplication and composition
(this is the case for the endomorphisms of any abelian group).
In particular, the endomorphisms and so the Dirichlet characters form a group w.r.t. pointwise multiplication.
\begin{definition}
    Denote by $\mathrm{Dir}(q)$ the set of Dirichlet characters modulo $q$.
\end{definition}
By Corollary~\ref{prop:dirichlet_character_image_group} each Dirichlet character resp. group endomorphism of $(\Z/q\Z)^\units$ is determined by its value at a primitive root of unity $r \in (\Z/q\Z)^\units$.
Further, by Corollary~\ref{prop:dirichlet_character_primitive_root}, we also have a Dirichlet character resp. group endomorphism for each possible value at a primitive root of unity.
Hence
\begin{equation*}
    |\mathrm{Dir}(q)| = |\mathrm{End}((\Z/q\Z)^\units)| = |(\Z/q\Z)^\units| = q - 1
\end{equation*}
It follows that there are exactly $q - 1$ distinct Dirichlet characters modulo a prime $q$.
\begin{remark}
    It is again a fact that $(\Z/p^k\Z)^\units$ is cyclic for an odd prime $p$ and $k \geq 1$.
    Hence, everything up to now can also be done for odd prime powers, if we replace $q - 1$ by $\phi(q)$.
\end{remark}
Because of Lemma~\ref{prop:dirichlet_series_at_one} it might seem like a good idea to study in which cases the value $\sum_{x \in (\Z/q\Z)^\units} \chi(x)$ is zero.
\begin{proposition}[Question (iii)]
    \label{prop:finite_sums_characters}
    Let $\chi_0$ be the trivial Dirichlet character given by $r \mapsto 1$.
    Then
    \begin{align*}
        \sum_{a \in (\Z/q\Z)^\units} \chi(a) &= \begin{cases}
            q - 1 & \text{if $\chi = \chi_0$} \\
            0 & \text{otherwise}
        \end{cases}, \\
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a) &= \begin{cases}
            q - 1 & \text{if $a \equiv 1 \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{align*}
    Furthermore, for $b \perp q$ have
    \begin{equation*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a)\overline{\chi(b)} = \begin{cases}
            q - 1 & \text{if $a \equiv b \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
\end{proposition}
\begin{proof}
    Clearly
    \begin{equation*}
        \sum_{a \in (\Z/q\Z)^\units} \chi_0(a) = q - 1 \quad \text{and} \quad \sum_{\chi \in \mathrm{Dir}(q)} \chi(1) = \sum_{\chi \in \mathrm{Dir}(q)} 1 = q - 1
    \end{equation*}
    So it is left to show that we get zero in the other cases.

    Consider a Dirichlet character $\chi \neq \chi_0$ given by $r \mapsto \xi$ for a primitive $q$-th root of unity $\xi \neq 1$.
    Then
    \begin{equation*}
        \sum_{a \in (\Z/q\Z)^\units} \chi(a) = \sum_{k = 0}^{q - 2} \chi(r^k) = \sum_{k = 0}^{q - 2} \xi^k = \frac {1 - \xi^{q - 1}} {q - \xi} = 0 
    \end{equation*}
    By using the earlier results on the structure of $\mathrm{Dir}(q)$ we see that for $a \equiv r^k \not\equiv 1 \pmod q$, have
    \begin{align*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a) &= \sum_{\chi \in \mathrm{Dir}(q)} \chi(r)^k \\
        &= \sum_{\xi \ \text{$q$-th root of unity}} \xi^k = \sum_{l = 0}^{q - 2} \omega^{kl} = \frac {1 - (\omega^{q - 1})^k} {1 - \omega^k} = 0 
    \end{align*}
    where $\omega \in \C$ is any primitive $q$-th root of unity and $r \in \Z$ is a primitive root modulo $q$.

    For the last part, note that for any $q$-th root of unity $\xi$, we have $\xi\overline{\xi} \in \R$ with $\xi\overline{\xi} = |\xi|^2 > 0$.
    Furthermore, $\overline{\xi}$ is also a $q$-th root of unity, and so we see that $\xi\overline{\xi} = 1$ (the only real, positive root of unity is 1).
    It follows that for any Dirichlet character $\chi$ have $\overline{\chi([a])} = \chi([a]^{-1})$.
    Thus
    \begin{equation*}
        \sum_{\chi \in \mathrm{Dir}(q)} \chi(a)\overline{\chi(b)} = \sum_{\chi \in \mathrm{Dir}(q)} \chi([a][b]^{-1}) = \begin{cases}
            q - 1 & \text{if $[a][b]^{-1} = 1 \in (\Z/q\Z)^\units$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    The condition $ab^{-1} = 1$ is equivalent to $a \equiv b \pmod q$, so the claim follows.
\end{proof}
Using these basic results, we can now prove facts on the Dirichlet series of characters.
\begin{proposition}[Question (iv)]
    \label{prop:representation_mangoldt_series_congruent_a}
    Let $a \perp q$. Then for $\Re(s) > 1$ have
    \begin{equation*}
        \sum_{n \equiv a \ (\text{mod} \ q)} \frac {\Lambda(n)} {n^s} = \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \frac {\Lambda(n) \chi(n)} {n^s}
    \end{equation*}
    (All those series obviously converge absolutely since $\Re(s) > 1$)
\end{proposition}
\begin{proof}
    By Prop.~\ref{prop:finite_sums_characters} we have for all $n \in \N$ that
    \begin{equation*}
        \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \chi(n) = \begin{cases}
            1 & \text{if $a \equiv n \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    It follows that
    \begin{align*}
        \sum_{n \equiv a \ (\text{mod} \ q)} \Lambda(n) n^{-s} =& \sum_{n \geq 1} \Lambda(n) n^{-s} \begin{cases}
            1 & \text{if $a \equiv n \pmod q$} \\
            0 & \text{otherwise}
        \end{cases} \\
        =& \sum_{n \geq 1} \Lambda(n) n^{-s} \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \chi(n) \\
        =& \frac 1 {q - 1} \sum_{\chi \in \mathrm{Dir}(q)} \overline{\chi(a)} \sum_{n \geq 1} \Lambda(n) \chi(n) n^{-s} 
    \end{align*}
    as infinite summation clearly commutes with finite sums.
\end{proof}
\begin{example}[Question (v)]
    \label{ex:nontrivial_dirichlet_characters_mod_5}
    We consider the Dirichlet characters mod 5.
    A (primitive) 5-th root of unity $\omega_5 \in \C$ is given by $\omega_5 = \exp(2\pi i/5)$.
    On the other hand, a primitive root modulo 5 is e.g. given by $r = 2$ since $2^2 \equiv -1 \pmod 5$.
    Thus we have the trivial Dirichlet character $\chi_0$ and $5 - 1 = 4$ nontrivial Dirichlet characters mod 5, namely those given by
    \begin{align*}
        \chi_1: 1 &\mapsto 1,\ 2 \mapsto \omega_5 = \exp(2\pi i/5),\ 3 \mapsto \omega_5^3 = \exp(6\pi i/5),\ 4 \mapsto \omega_5^2 = \exp(4\pi i/5),\ \\
        \chi_2: 1 &\mapsto 1,\ 2 \mapsto \omega_5^2 = \exp(4\pi i/5),\ 3 \mapsto \omega_5 = \exp(2\pi i/5),\ 4 \mapsto \omega_5^4 = \exp(8\pi i/5),\ \\
        \chi_3: 1 &\mapsto 1,\ 2 \mapsto \omega_5^3 = \exp(6\pi i/5),\ 3 \mapsto \omega_5^4 = \exp(8\pi i/5),\ 4 \mapsto \omega_5^1 = \exp(2\pi i/5),\ \\
        \chi_4: 1 &\mapsto 1,\ 2 \mapsto \omega_5^4 = \exp(8\pi i/5),\ 3 \mapsto \omega_5^2 = \exp(4\pi i/5),\ 4 \mapsto \omega_5^3 = \exp(6\pi i/5),\ \\
    \end{align*}
\end{example}

@@

    More concretely, consider
    \begin{equation*}
        f: \N \to \C, \quad n \mapsto \begin{cases}
            1 & \text{if $n \equiv a \pmod q$} \\
            0 & \text{otherwise}
        \end{cases}
    \end{equation*}
    Note that this function is not multiplicative.
    After applying the first step, we are interested in the poles of the meromorphic function
    \begin{equation*}
        \sum_{n \geq 1} f(n) \Lambda(n) n^{-s}
    \end{equation*}
    I will not prove this here, but via partial summation, we can find a meromorphic continuation to $\Re(s) > 0$.

    However, here we get stuck.
    It might seem natural to consider the function
    \begin{equation*}
        \zeta_a(s) := \sum_{n \geq 1} f(n) n^{-s}
    \end{equation*}
    which also has a meromorphic continuation to $\Re(s) > 0$.
    Its derivative is clearly given by
    \begin{equation*}
        \zeta_a'(s) = \sum_{n \geq 1} f(n) \log(n) n^{-s}
    \end{equation*}
    To continue, we might try to consider
    \begin{equation*}
        \zeta_a'(s)\Bigl( \sum_{n \geq 1} \mu(n) f(n) n^{-s} \Bigr) = \sum_{n \geq 1} \frac {(f\log * f\mu)(n)} {n^s}
    \end{equation*}
    But since $f$ has no nice compatibility with multiplication or $*$, it is not easy to make something of this.
    In particular
    \begin{equation*}
        f\Lambda \neq f\log * f\mu \quad \text{and} \quad \frac 1 {\zeta_a(s)} \neq \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    Note that we find $n \equiv a \pmod q$ such that $n$ has no nontrivial factor $\equiv a \pmod q$.
    E.g. take $n = 9 \equiv 4 \pmod 5$, then
    \begin{equation*}
        \log(3) = (f\Lambda)(6) \neq \sum_{cd = 9} f(c)\log(c)f(d)\mu(d) = 0
    \end{equation*}
    Furthermore, the equality
    \begin{equation*}
        \frac 1 {\zeta_a(s)} = \sum_{n \geq 1} f(n) \mu(n) n^{-s}
    \end{equation*}
    would imply
    \begin{equation*}
        1 = \zeta_a(s) \sum_{n \geq 1} f(n) \mu(n) n^{-s} = \sum_{n \geq 1} \frac {(f * f\mu)(n)} {n^s}
    \end{equation*}
    which in turns implies that $f * f\mu = \delta$.
    However, we clearly find counterexamples for that, e.g. for $q = 5, a = 4$ find $(f * f\mu)(1) = 0$.
    The underlying reason is that the divisors of $n$ may lie in a different congruence class modulo $q$ than $n$.
